Tabular Constraint Learning

### 2016-04-14 Paper/ECAI/tabular.tex
merged mostly

\tias{Spreadsheet is an instance of tabular data, but not the only one?}

\tias{Rewrite}
\section{Introduction}
Millions of people across the world use spreadsheets everyday.
Yet, many people lack a complete understanding of the structures and existing dependencies in the data.
Indeed, the most useful part of spreadsheets is often the formulas and often these formula disappear when a file is exported and shared, for example as a CSV file.
This brings us to the question: can we discover or reconstruct structural relations in flat tabular spreadsheet data?
Moreover, can we accomplish this task in a general way that allows declarative specification of constraints to discover?

\samuel{Maybe slightly more formal:}
Let us look at an example in Figure~\ref{fig:main_example}, the header names already suggest the usage of standard spreadsheet operations such as \textit{average} or \textit{sum}.
If we look at Table~1 and at the first row in Table~2, it is clear that one could have selected the column called \textbf{1st Quarter}, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns.
Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2.
I.e., all operations are vectorized and make use of the standard spreadsheet operations.

Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
\samuel{Maybe slightly more formal:}
That gives us already a hint how this problem is different from the standard data mining settings, when the data is in rows and variables are in columns. Here everything is mixed, some values are missing in the results, some missing the arguments, even a header might have missing values. The data is relational on one hand, since we have multiple table with relationships between them. And on the other hand, the data is numeric, since people tend to use spreadsheets for financial and accounting computations, which is heavily numeric.

Given these novel aspects, we conclude that understanding and learning constraints in tabular data is a new and challenging problem setting.
Futhermore, it is not only challenging for a machine but for a human as well.
\samuel{Maybe slightly more formal:} Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the Spreadsheet risk\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaw in the famous economics papers \cite{flaw_excel} and caused billion-losses in the financial industry. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

We consider various sets of cells in tables:
\begin{itemize}
  \item {\bf Cells}: Each entry in the matrix/table is called  a \textit{cell}.
  We use the notation $T[a,b]$ to refer to a particular cell in the table~$T$ at the row with index~$a$ and the column with index~$b$.
  \item {\bf Columns and rows}: To refer to the $a$-th row (column) we use the following syntax $T[a,{:}]$ ($T[{:},a]$), to refer to a subrange from $a$ to $b$ we use $T[a{:}b,]$ ($T[,a{:}b]$).
  \item
  A \textit{vector} is either a column or a row that is type-consistent.
  If a vector is a row (column), we say that it has a \textit{row} (\textit{column}) orientation.
  \item
  A \textit{group} is a subrange of vectors with the same orientation in a table.
  We use the following notation to refer to a row (column) group $G$ in a table $T$ with rows (columns) ranging from $a$ to $b$: $G = T[a{:}b,:]$ ($G = T[{:},a{:}b]$), where $a,b$ are natural numbers.
  We denote as the \textit{length} of a row (column) group $G$, written as \textit{length(G)}, the number of its columns (rows). We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if \luc{all?????} its vectors contain numeric (textual, etc) elements.
  \item
  A \textit{subgroup} $G'$ is a subrange of vectors belonging to a group $G$. If the subgroup $\{v\}$ must contain only one vector, we write $v \in G$.
\end{itemize}
For notational convenience, we will refer to cells with $C$,  to vectors with $V$, and to groups and subgroups with $G$.
When we do not want to distinguish between cells, columns, rows, vectors or groups, we shall talk about a {\em set of cells}.
\luc{add an example of group/ subgroup}

\tias{Use generic constraint}
A \textit{constraint} definition is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them.
\begin{itemize}
\item
\textit{\CName} is the name of the constraint together with the name of the groups,
\item \textit{\CSignature} is specifying a set of restrictions on the types and variables of the constraint,
% the properties of the group assignments, corresponding to the variables in \CName, such as their types,
e.g., requiring the variables to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are restrictions on the group meta-information not on the actual group content.
\item \textit{\CFunction} defines the actual constraint that needs to be satisfied in the data.
\end{itemize}

\paragraph{Example}
The constraint \textit{rank} has
\begin{itemize}
\item \CName = \textit{$g_Y$ = RANK($g_X$)}, where $g_X$ and $g_Y$ are the group variables;
\item \CSignature = subgroup $g_X$ (associated with X) is numeric, $g_Y$ is integer and the length of vectors in $g_X$ is the same as in $g_Y$; and
\item
\CFunction = \textit{$g_Y$ = RANK($g_X$)} iff $g_X \in G_X, g_Y \in G_Y$ and each value in $g_X$ has the rank (possibly with ties) specified in $g_Y$.
\end{itemize}

\samuel{2nd problem statement (from approach), remove EQUAL}

@@

%A \textit{tabular solution} $S$ to a spreadsheet constraint $C$ (with $n$ variables) is a subgroup assignment $(g_1, \dots, g_n)$ where $g_i$, such that $(g_1,\dots,g_n)$ satisfies both the \CSignature and \CFunction of the constraint $C$. \sergey{to finish this definition later, need to say that $g_i$ is the subgroup assigned to the $i$-th variable in $C$}

>>>>

\section{Introduction}
Millions of people across the world use spreadsheets everyday.
Yet, many people lack a complete understanding of the structures and existing dependencies in the data.
Indeed, the most useful part of spreadsheets is often the formulas and often these formula disappear when a file is exported and shared, for example as a CSV file.
This brings us to the question: can we discover or reconstruct structural relations in flat tabular spreadsheet data?
Moreover, can we accomplish this task in a general way that allows declarative specification of constraints to discover?

Let us look at an example in Figure~\ref{fig:main_example}, the header names already suggest the usage of standard spreadsheet operations such as \textit{average} or \textit{sum}.
If we look at Table~1 and at the first row in Table~2, it is clear that one could have selected the column called \textbf{1st Quarter}, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns.
Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2.
I.e., all operations are vectorized and make use of the standard spreadsheet operations.

Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.

That gives us already a hint how this problem is different from the standard data mining settings, when the data is in rows and variables are in columns. Here everything is mixed, some values are missing in the results, some missing the arguments, even a header might have missing values. The data is relational on one hand, since we have multiple table with relationships between them. And on the other hand, the data is numeric, since people tend to use spreadsheets for financial and accounting computations, which is heavily numeric.

Given these novel aspects, we conclude that understanding and learning constraints in tabular data is a new and challenging problem setting.
Futhermore, it is not only challenging for a machine but for a human as well.

Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the Spreadsheet risk\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaw in the famous economics papers \cite{flaw_excel} and caused billion-losses in the financial industry \cite{spreadsheet_risk_loss}. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

We consider various sets of cells in tables:
\begin{itemize}
  \item {\bf Cells}: Each entry in the matrix/table is called  a \textit{cell}.
  We use the notation $T[a,b]$ to refer to a particular cell in the table~$T$ at the row with index~$a$ and the column with index~$b$.
  \item {\bf Columns and rows}: To refer to the $a$-th row (column) we use the following syntax $T[a,{:}]$ ($T[{:},a]$), to refer to a subrange from $a$ to $b$ we use $T[a{:}b,]$ ($T[,a{:}b]$).
  \item
  A \textit{vector} is either a column or a row that is type-consistent.
  If a vector is a row (column), we say that it has a \textit{row} (\textit{column}) orientation.
  \item
  A \textit{group} is a subrange of vectors with the same orientation in a table.
  We use the following notation to refer to a row (column) group $G$ in a table $T$ with rows (columns) ranging from $a$ to $b$: $G = T[a{:}b,:]$ ($G = T[{:},a{:}b]$), where $a,b$ are natural numbers.
  We denote as the \textit{length} of a row (column) group $G$, written as \textit{length(G)}, the number of its columns (rows). We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if \luc{all?????} its vectors contain numeric (textual, etc) elements.
  \item
 A \textit{subgroup} $g$ is a subrange of vectors belonging to a group $G$. If the subgroup $g$ must contain only one vector, we write $g \in G$.
%We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If a subgroup $g$ must contain only one vector, we write $g \in G$.
% A \textit{subgroup} $G'$ is a subrange of vectors belonging to a group $G$. If the subgroup $\{v\}$ must contain only one vector, we write $v \in G$.
\end{itemize}
For notational convenience, we will refer to cells with $C$ \sergey{maybe better name? confusing with constaraints},  to vectors with $V$, and to groups and subgroups with $G$.

When we do not want to distinguish between cells, columns, rows, vectors or groups, we shall talk about a {\em set of cells}.
\luc{add an example of group/ subgroup}

A \textit{Spreadsheet constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. 
\begin{itemize}
\item
\textit{\CName} is the textual name of the constraint together with the list of its variables $v_1,\dots,v_n$.
\item
  \textit{\CSignature} is a constraint specifying the properties of the group assignments, corresponding to the variables $v_1,\dots,v_n$ in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. 
\item \textit{\CFunction} is a constraint specifying that the data in the subgroups satisfies the function.

\paragraph{Example}
\sergey{need to use this below:}

  For example, a spreadsheet constraint \textit{rank} has \CName \textit{$v_y$ = RANK($v_x$)}, where $v_x$ and $v_y$ are the variables; its \CSignature is: the subgroup assigned to $v_x$ must be numeric, $v_y$ must be integer and the length of vectors in $v_x$ is the same as in $v_y$; its \CFunction is the following constraint: the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ is a solution iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_x$ has the rank (possibly with ties) specified in $g_y$.

The constraint \textit{rank} has
\begin{itemize}
\item \CName = \textit{$g_Y$ = RANK($g_X$)}, where $g_X$ and $g_Y$ are the group variables;
\item \CSignature = subgroup $g_X$ (associated with X) is numeric, $g_Y$ is integer and the length of vectors in $g_X$ is the same as in $g_Y$; and
\item
\CFunction = \textit{$g_Y$ = RANK($g_X$)} iff $g_X \in G_X, g_Y \in G_Y$ and each value in $g_X$ has the rank (possibly with ties) specified in $g_Y$.
\end{itemize}

@@

Let us now define what is a tabular solution for a spreadsheet constraint in general. Let $C$ denote a spreadsheet constraint, $S$ a mapping from variables to the subgroups of \groups, and $C^S$ the constraint $C$, in which all variables are substituted with the corresponding subgroups from $S$, then $S$ is a \textit{tabular solution} to $C$ iff both $\CSignature^S$ and $\CFunction^S$ hold.
